---
title: "Correlation Analysis"
author: "Mae Rennick"
date: "2023-08-28"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: yeti
    highlight: tango
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r setup, include=FALSE, warning= FALSE, message= FALSE}
knitr::opts_chunk$set(echo = TRUE, warning= FALSE, message= FALSE)

# Packages ----
library(tidyverse)    # For data manipulation, plotting, etc. 
library(here)    # For reproducible data importation 
library(janitor)      # For text data manipulation
library(d3heatmap)
library(ggrepel)


```
 
The analysis conducted in the R notebook, inspired by the work of Toumi (2018), delves into the realm of data visualization using the R programming language. Drawing from the insights shared in Toumi's work on "R for data visualization," the notebook likely encompasses a comprehensive exploration of various techniques and libraries available in R for creating impactful and informative visual representations of data. Whether focusing on basic charts or advanced graphical displays, the analysis likely showcases the versatility of R's visualization capabilities. With the provided citation serving as a foundational reference, the notebook is expected to empower data analysts and researchers with the tools to effectively communicate their findings through compelling visuals, ultimately aiding in clearer understanding and decision-making processes. 

```{r}



###Import data
teams <- read_csv(here("MIT_sloan_data", "nhl_teams_data_mit.csv"))

###Preview teams
summary(teams)

team_names<- first_column <- teams[, 1] %>% 
  rename(team_name= Team) %>% 
  mutate(Team= row_number())

###Prepare dataset
row.names(teams) <- teams$Team #Set row names to be teams
teams$Season <- NULL #Remove seasons column
teams$Team <- NULL #Remove teams column
teams$GP <- NULL #Remove GP
teams$TOI <- NULL #Remove TOI


###Rename column names
names(teams) <- c("CF%", "CF/60", "CA/60","GF%", "GF/60", 
                  "GA/60", "xGF%", "xGF/60", "xGA/60", 
                  "Sh%", "Sv%", "PDO")

###View objects stored in teams dataset
names(teams) 

###Set dataframe to a matrix
team_matrix <- data.matrix(teams)

###Make interactive heatmap
heatmap <- d3heatmap(team_matrix, colors = "Blues", scale = "col",
                     dendrogram = "row", k_row = 4)

heatmap


#########################PCA CLUSTERING#######################################



###Perform PCA on teams data
pca_teams <- princomp(teams, cor = TRUE) 

###View objects stored in pcaCars
names(pca_teams)

###Proportion of variance explained
summary(pca_teams)

###Plot scree plot
plot(pca_teams, type = "l")

###Cluster teams
teamsHC <- hclust(dist(pca_teams$scores), method = "ward.D2")

###Plot dendrogram
plot(teamsHC)

###Cut dendogram into 5 clusters
teams_clusters <- cutree(teamsHC, k = 5)

###Add clusters to dataframe of scores
teams_df <- data.frame(pca_teams$scores, "cluster" = factor(teams_clusters))
teams_df <- transform(teams_df, cluster_name = paste("Cluster",teams_clusters))

###Plot
pcaplot <- ggplot(teams_df,aes(x=Comp.1, y=Comp.2)) +
  theme_minimal() +
  geom_hline(yintercept = 0, color = "gray70") +
  geom_vline(xintercept = 0, color = "gray70") +
  geom_point(aes(color = cluster), alpha = 0.55, size = 3) +
  xlim(-5, 6) + 
  labs(x = "PC1", y ="PC2", 
       title="PCA Clusters from Hierarchichal Clustering of NHL Teams", 
       subtitle="Data: corsica.hockey")

###View plot
pcaplot

team_name_data <- read_csv(here("MIT_sloan_data", "nhl_teams_data_mit.csv"))

row.names(team_name_data) <- team_name_data$Team #Set row names to be teams

###Add labels
pcaplot + geom_text(aes(y = Comp.2 + 0.25, label = rownames(team_name_data)))

###Use ggrepel to repel overlapping text labels 
pcaplot + geom_text_repel(aes(y = Comp.2 + 0.009, label = rownames(team_name_data))) 


```

