---
title: "DAY 2: PCA analysis"
author: "Mae Rennick"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: yeti
    highlight: tango
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r setup, include=FALSE, warning= FALSE, message= FALSE}
knitr::opts_chunk$set(echo = TRUE, warning= FALSE, message= FALSE)

# Packages ----
library(tidyverse)    # For data manipulation, plotting, etc. 
library(here)    # For reproducible data importation 
library(janitor)      # For text data manipulation
library(d3heatmap)
library(ggrepel)


```
 
The analysis conducted in the R notebook, inspired by the work of Toumi (2018), delves into the realm of data visualization using the R programming language. Drawing from the insights shared in Toumi's work on "R for data visualization," the notebook likely encompasses a comprehensive exploration of various techniques and libraries available in R for creating impactful and informative visual representations of data. Whether focusing on basic charts or advanced graphical displays, the analysis likely showcases the versatility of R's visualization capabilities. With the provided citation serving as a foundational reference, the notebook is expected to empower data analysts and researchers with the tools to effectively communicate their findings through compelling visuals, ultimately aiding in clearer understanding and decision-making processes. 

### Heatmapping

Statistics analyzed: 

1. **CF% (Corsi For Percentage):** the percentage of shot attempts (shots on goal, missed shots, and blocked shots) taken by a team while a particular player or team is on the ice. A high CF% indicates that a player or team is controlling play and generating more shot attempts than they are giving up.

2. **CF/60 (Corsi For per 60 minutes):** the average number of shot attempts (shots on goal, missed shots, and blocked shots) a player or team generates per 60 minutes of ice time.

3. **CA/60 (Corsi Against per 60 minutes):** the average number of shot attempts (shots on goal, missed shots, and blocked shots) a player or team gives up per 60 minutes of ice time.

4. **GF% (Goals For Percentage):** the percentage of goals scored by a team while a particular player or team is on the ice. A high GF% indicates that a player or team is contributing to more goals being scored compared to the goals they are conceding.

5. **GF/60 (Goals For per 60 minutes):** the average number of goals scored by a player or team per 60 minutes of ice time.

6. **GA/60 (Goals Against per 60 minutes):** the average number of goals conceded by a player or team per 60 minutes of ice time.

7. **xGF% (Expected Goals For Percentage):** the probability of a team scoring goals based on the quality of shot attempts (location, type of shot, etc.) generated while a particular player or team is on the ice.

8. **xGF/60 (Expected Goals For per 60 minutes):** the average expected number of goals a player or team would score per 60 minutes of ice time based on the quality of shot attempts.

9. **xGA/60 (Expected Goals Against per 60 minutes):** the average expected number of goals a player or team would concede per 60 minutes of ice time based on the quality of shot attempts against.

10. **Sh% (Shooting Percentage):** the percentage of shots on goal that result in goals. It measures the efficiency of a player or team's shooting.

11. **Sv% (Save Percentage):**  percentage of shots on goal that a goaltender stops, indicating their efficiency in preventing goals.

12. **PDO:** a combination of a team's shooting percentage (Sh%) and its goaltender's save percentage (Sv%). It's often used as a luck indicator, as extremely high or low PDO values tend to regress toward the mean over time.


```{r}

###Import data
teams <- read_csv(here("MIT_sloan_data", "nhl_teams_data_mit.csv"))

###Preview teams
summary(teams)

team_names<- first_column <- teams[, 1] %>% 
  rename(team_name= Team) %>% 
  mutate(Team= row_number())

###Prepare dataset
row.names(teams) <- teams$Team #Set row names to be teams
teams$Season <- NULL #Remove seasons column
teams$Team <- NULL #Remove teams column
teams$GP <- NULL #Remove GP
teams$TOI <- NULL #Remove TOI


###Rename column names
names(teams) <- c("CF%", "CF/60", "CA/60","GF%", "GF/60", 
                  "GA/60", "xGF%", "xGF/60", "xGA/60", 
                  "Sh%", "Sv%", "PDO")

###View objects stored in teams dataset
names(teams) 

###Set dataframe to a matrix
team_matrix <- data.matrix(teams)

###Make interactive heatmap
heatmap <- d3heatmap(team_matrix, colors = "Blues", scale = "col",
                     dendrogram = "row", k_row = 4)

heatmap



```



### PCA Analysis

```{r}
#########################PCA CLUSTERING#######################################



###Perform PCA on teams data
pca_teams <- princomp(teams, cor = TRUE) 

###View objects stored in pcaCars
names(pca_teams)

###Proportion of variance explained
summary(pca_teams)

###Plot scree plot
plot(pca_teams, type = "l")

###Cluster teams
teamsHC <- hclust(dist(pca_teams$scores), method = "ward.D2")

###Plot dendrogram
plot(teamsHC)

###Cut dendogram into 5 clusters
teams_clusters <- cutree(teamsHC, k = 5)

###Add clusters to dataframe of scores
teams_df <- data.frame(pca_teams$scores, "cluster" = factor(teams_clusters))
teams_df <- transform(teams_df, cluster_name = paste("Cluster",teams_clusters))

###Plot
pcaplot <- ggplot(teams_df,aes(x=Comp.1, y=Comp.2)) +
  theme_minimal() +
  geom_hline(yintercept = 0, color = "gray70") +
  geom_vline(xintercept = 0, color = "gray70") +
  geom_point(aes(color = cluster), alpha = 0.55, size = 3) +
  xlim(-5, 6) + 
  labs(x = "PC1", y ="PC2", 
       title="PCA Clusters from Hierarchichal Clustering of NHL Teams", 
       subtitle="Data: corsica.hockey")

###View plot
pcaplot

team_name_data <- read_csv(here("MIT_sloan_data", "nhl_teams_data_mit.csv"))

row.names(team_name_data) <- team_name_data$Team #Set row names to be teams

###Add labels
pcaplot + geom_text(aes(y = Comp.2 + 0.25, label = rownames(team_name_data)))

###Use ggrepel to repel overlapping text labels 
pcaplot + geom_text_repel(aes(y = Comp.2 + 0.009, label = rownames(team_name_data))) 


```

```{r}
# ... (PCA analysis and clustering)

# Scree Plot with labels
plot(pca_teams, type = "l", main = "Scree Plot - Explained Variance")
text(1:ncol(teams), pca_teams$sdev^2, labels = round(pca_teams$sdev^2, 2), pos = 3)

# Enhanced Dendrogram
plot(teamsHC, main = "Dendrogram - Team Clusters")

# Enhanced PCA Scatter Plot
pcaplot <- ggplot(teams_df, aes(x = Comp.1, y = Comp.2, color = cluster)) +
  theme_minimal() +
  geom_hline(yintercept = 0, color = "gray70") +
  geom_vline(xintercept = 0, color = "gray70") +
  geom_point(alpha = 0.7, size = 3) +
  xlim(-5, 6) +
  labs(x = "PC1", y = "PC2", 
       title = "PCA Clusters from Hierarchical Clustering of NHL Teams", 
       subtitle = "Data: corsica.hockey")

# Add team labels with ggrepel to prevent overlap
library(ggrepel)
pcaplot <- pcaplot + geom_text_repel(aes(label = rownames(team_name_data)))

pcaplot

```



The teams within each cluster exhibit similar performance characteristics.


The scree plot indicates that the first two principal components capture a significant portion of the variance in the data. This suggests that a two-dimensional representation is informative for distinguishing between teams based on these metrics.

```{r}

### Team Labels on Dendrogram

# Create a list of labels with team names
label_names <- rownames(team_name_data)

# Plot dendrogram with labeled leaves
plot(teamsHC, main = "Dendrogram with Labeled Leaves", labels = label_names)


# Calculate mean metrics for each cluster
cluster_mean_metrics <- teams_df %>%
  group_by(cluster) %>%
  summarize_all(mean)

# Melt the data for easier plotting
cluster_mean_metrics_melted <- melt(cluster_mean_metrics, id.vars = "cluster", variable.name = "metric", value.name = "mean_value")

# Create a bar plot showing mean metrics for each cluster
cluster_metrics_plot <- ggplot(cluster_mean_metrics_melted, aes(x = metric, y = mean_value, fill = factor(cluster))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  labs(x = "Metrics", y = "Mean Value", title = "Mean Metrics for Each Cluster") +
  theme_minimal() +
  theme(legend.position = "top")

cluster_metrics_plot

# Calculate within-cluster variability using the standard deviation of each principal component within each cluster
within_cluster_variability <- teams_df %>%
  group_by(cluster) %>%
  summarize(sd_comp1 = sd(Comp.1), sd_comp2 = sd(Comp.2))

# Create a scatter plot to visualize within-cluster variability
within_cluster_variability_plot <- ggplot(within_cluster_variability, aes(x = sd_comp1, y = sd_comp2, color = factor(cluster))) +
  geom_point(size = 3, alpha = 0.7) +
  labs(x = "Standard Deviation of PC1", y = "Standard Deviation of PC2", title = "Within-Cluster Variability in PCA Space") +
  theme_minimal()

within_cluster_variability_plot


```

